Effectful Semantics for Substructural SSA

### 2023-11-09 pldi24/paper.tex
Figure

\usepackage{listings}
\usepackage{xcolor}

@@

In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks of the following form:
\begin{lstlisting}
'label:
  var = instruction
  ...
  var = instruction
  terminator
\end{lstlisting}
A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
\begin{lstlisting}
'entry:
  x = 7
  x = add x 2
  ret x
\end{lstlisting}
is not in SSA form, but can easily be converted into SSA form by rewriting it as
\begin{lstlisting}
'entry:
  x0 = 7
  x1 = add x0 2
  ret x1
\end{lstlisting}
Unfortunately, things get a little more tricky when we try to define variables in loops or branches. For example, consider the following code:
\begin{lstlisting}
'entry:
  br arg0 'left `right
'left:
  x = 7
  jmp 'exit
'right:
  x = 3
  jmp 'exit
'exit:
  ret x
\end{lstlisting}
There are two competing definitions of \(x\) in the left and right blocks, and it’s unclear how to disambiguate between them. The usual solution is to introduce a \(\phi\)-node. In particular, we can write:
\begin{lstlisting}
'entry:
  br arg0 'left `right
'left:
  x0 = 7
  jmp 'exit
'right:
  x1 = 3
  jmp 'exit
'exit:
  x2 = phi 'left x0 'right x1
  ret x2
\end{lstlisting}
\TODO{make above into graphs?}
\TODO{segue}
The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.

>>>>

\usepackage{listings}
\usepackage{xcolor}
\usepackage{subfig}

@@

In a standard SSA-based compiler, to represent a function, we begin by decomposing it into basic blocks, which can be viewed as \textit{label}, followed by a sequence of \textit{instructions}, each defining a \textit{variable}, followed by a \textit{terminator}.
A program written in this manner being in ``SSA form'' simply means that every variable name is defined exactly once. For example,
\TODO{text}

\newsavebox\notssafib

\begin{lrbox}{\notssafib}
\begin{minipage}{0.3\textwidth}
\begin{lstlisting}[frame=none,numbers=none]
'entry:
  m = 0
  n = 1
  brz i 'exit 'loop
'loop:
  t = add m n
  m = n
  n = t
  N = sub i 1
  brz i 'exit 'loop
'exit:
  ret m
\end{lstlisting}
\end{minipage}
\end{lrbox}

\newsavebox\ssafib

\begin{lrbox}{\ssafib}
\begin{minipage}{0.3\textwidth}
\begin{lstlisting}[frame=none,numbers=none]
'entry:
  m0 = 0
  n0 = 1
  brz N 
    'exit (m0) 
    'loop (i, m0, n0)
'loop(i0, m1, n1):
  m2 = n1
  n2 = add m1 n1
  i1 = sub i0 1
  brz z1 
    'exit (m2) 
    'loop (i1, m2, n2)
'exit(m3):
  ret m3
\end{lstlisting}
\end{minipage}
\end{lrbox}

\newsavebox\ourfib

\begin{lrbox}{\ourfib}
\begin{minipage}{0.3\textwidth}
\begin{lstlisting}[frame=none,numbers=none]
let m0 = 0;
let n1 = 1;
if eq i 0 {
  br 'exit(m0)
} else {
  br 'loop(i, m0, n0)
} where
'loop(i0, m1, n1) =>
  let m2 = n1;
  let n2 = add m1 n1;
  let i1 = sub i0 1;
  if eq i1 0 {
    br 'exit(m2)
  } else {
    br 'loop(i1, m2, n2)
  }
'exit(m3) => ret m3
\end{lstlisting}
\end{minipage}
\end{lrbox}

\begin{figure}
  \centering
  \subfloat[3-address code]{\usebox\notssafib}
  \quad
  \subfloat[Basic blocks with arguments]{\usebox\ssafib}
  \quad
  \subfloat[Our syntax]{\usebox\ourfib}
  \caption{A simple program for computing the \texttt{i}th Fibonacci number, represented as 3-address code, basic-blocks with arguments, and in our SSA syntax}
  \Description{}
  \label{fig:ssa-examples}
\end{figure}
